plantecophys - an R package for analysing and modelling leaf gas exchange data

### 2015-07-07 manuscript.Rmd
Formatting requirements for PLOS ONE; reference fixes

@@

@@

# Set Flag to include figures in text or not.
includeFigs <- TRUE

@@

# Find current SHA of plantecophys
pkgrep <- repository("c:/repos/plantecophys")
r <- reflog(pkgrep, verbose=FALSE)
cur_sha <- substr(r[[1]]@sha, 1, 7)
```

@@

@@

Since the advent of portable gas exchange instruments, a wealth of data on leaf gas exchange of CO~2~ and H~2~O has been collected  [@long2003]. These data play a central role in physiological plant ecology [@pearcy1989]. Both photosynthesis and transpiration vary widely between plant species, and fluctuate rapidly in response to environmental drivers such as light, humidity and temperature. Not only do leaf gas exchange data allow detailed studies of the underlying plant physiology, they are also used to parameterize an important component of land surface models used to predict global water and carbon cycling [e.g. @bonan2014; @dekauwe2015].

@@

(1)
$$A_n = min(A_c, A_j) - R_d$$

@@

(2)
$$C_c = C_i - A_n/g_m$$

@@

\
Fitting the model to data requires some finesse because the model is not a single non-linear equation, but instead a minimum function of two non-linear equations that is sometimes difficult to fit. Moreover, sample sizes collected are often small due to time constraints in the field. A widely used published method requires the user to specify the transition of Vcmax to Jmax limitation [@sharkey2007], a process that is both both arbitrary and prevents batch analysis. Another method [@leafweb] requires online submission of data and fits the model without much control or knowledge of the fitting process [@gu2010]. Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see available online tools [@dekauwefitfarquhar; @landfluxtools]).

@@

(3)
$$ A_n = g_s(C_a - C_i) $$

\
where g~s~ is the conductance to CO~2~. We now have two equations for the photosynthesis rate. The 'demand function' is given by Eq. 1, and the 'supply function' in Eq. 2. At steady state these two equation should be equal, which can be graphically shown as in Fig. 1 (cf. [@farquhar1982]).

\
Because g~s~ itself responds to environmental drivers, we must then estimate this dependence with another model. The most successful, though empirical, g~s~ models is the Ball-Berry [@ballberry1987] class of models. This model posits an entirely empirical equation that describes the response of g~s~ to air humidity, CO~2~ and the rate of photosynthesis. This way, effects of leaf temperature and PPFD - both of which are known to affect g~s~ - are modelled through the dependency of A~n~ on these drivers. Thus the third equation is,

(4)
$$ g_s = g_0 + g_1 \frac{A_n}{C_a}f(D)$$

\
where D is the vapour pressure deficit (kPa), g~0~ and g~1~ are empirical parameters. An alternative approach to modelling g~s~ is through the hypothesis that stomata act optimally in the sense that they maximize photosynthesis while minimizing water loss. This hypothesis was first developed by Cowan and Farquhar [@cowan1977] and has seen many applications. Medlyn et al. [@medlyn2011] showed that the optimality hypothesis, when coupled to the FvC model, leads to an expression analogous to the Ball-Berry type models (Eq. 3), but with a different D response function (f(D) in Eq. 3) than the original Ball-Berry model, or yet another version introduced by Leuning [@leuning1995].

\
Finally we can combine the biochemical demand function of photosynthesis (Eq. 1) with the supply function (Eq. 2) and an expression for the dependency of g~s~ on environmental drivers (Eq. 3). This 'coupled' leaf gas exchange model [@leuning1990; @collatz1991] is implemented in many process-based ecosystem models [@wang1998; @bonan2014]. This model allows prediction of A~n~, g~s~ and leaf transpiration rate in response to all major environmental drivers (except soil water limitation), and key leaf traits (g~1~, V~cmax~, J~max~, R~d~).

\
Despite the widespread use of the FvCB model and the coupled leaf gas exchange model, tools to analyze the data are scattered and subject to little standardization. Fitting the FvCB model to CO~2~ response curves is a standard procedure but different methods can yield different parameter values, making comparisons difficult. The coupled leaf gas exchange model is not straightforward to implement, and I do not know of any standalone open-source implementations. I here describe the plantecophys package, implemented in the R language [@rfoundation]. The code is freely available (without restrictions), and managed with a version control system. The package is the results of our work on leaf and canopy modelling of photosynthesis and stomatal conductance [@medlyn2011; @barton2012; @peltoniemi2012; @medlyn2013; @duursma2013; @duursma2014], with many additions based on user requests.

@@

The 'plantecophys' package is implemented in native R, and has no dependencies on other packages. As such it builds easily, and is highly portable. The source code is maintained with git version control, and is hosted in an online repository (http://www.bitbucket.org/remkoduursma/plantecophys). The repository includes an issue tracker, where users can suggest changes or report bugs.  This paper describes version `r packageVersion("plantecophys")` (git SHA `r cur_sha`).

@@

### Fitting A-Ci curves

The `fitaci` function fits the FvCB model, yielding estimates of V~cmax~, J~max~ and R~d~ and their standard errors. Optionally R~d~ can be provided as a known value, otherwise it is estimated from the A-Ci curve. The user does not have to provide the transition point (see Fig. 1), as this is estimated by `fitaci` automatically. It is however an option to fix the transition point, which may be helpful to check whether the best fit was achieved. Standard errors are estimated with the usual methods for non-linear regression (`nls` function in base R). Instead of fitting the minimum function (Eq. 1), `fitaci~ uses the hyperbolic minimum, which avoids a discontinuity at the transition from A~c~ to A~j~ limitation. Finally, the user can provide an estimate of mesophyll conductance (g~m~) [following @ethier2004], in which case the fitted values of V~cmax~ and J~max~ can be interpreted as chloroplastic rates.

@@

\
The output of the above example is shown in Fig. 2. Additionally, `fitacis` can be sued to fit many curves at once, for example one for each species or site in a dataset. The function `Aci` simulates the A-Ci curve, that is, it returns A~n~ when C~i~ is known.

@@

The intersection of the supply and demand curves of photosynthesis (Fig. 1b) gives the operating intercellular CO~2~ concentration. This is solved by the `Photosyn` function. This flexible interface can be used to either 1) estimate A~n~ when C~i~ is known (`Photosyn(Ci=...)`, equivalent to `Aci(...)`), 2) estimate A~n~ when g~s~ is known (`Photosyn(GS=...)`) (cf. Fig. 1b) or c) solve C~i~ from the coupled leaf gas exchange model (Eqs. 1,3,4).

To demonstrate the use of the coupled gas exchange model, I visualize the temperature response of A~n~ when both T~leaf~ and D are varying. In field conditions, D is always strongly positively related to T~leaf~. The consequence is that when studying D or T~leaf~ responses in the field, both drivers have to be accounted for [@lin2012; @duursma2014]. In the FvCB model, J~max~, V~cmax~ and leaf respiration (R~d~) all depend on T~leaf~. The `Photosyn` function incorporates standard temperature sensitivities for all parameters of the FvCB model [following @medlyn2002]. Figure 3a shows simulated A-C~i~ curves and the solutions of the coupled leaf gas exchange models at a range of T~leaf~ and corresponding D [calculated following @duursma2014]. Figure 3b shows the output from `Photosyn`, which can be created with the following command. Note that in this example the default values of many parameters are used in the call to `Photosyn`, but all of these can be changed by the user.

@@

The `Photosyn` function assumes that the boundary layer conductance (g~bl~) is high compared to g~s~, so that T~leaf~ is close to T~air~. As an alternative, the `PhotosynEB` function calculates T~leaf~ from the leaf energy balance. Transpiration is calculated with the Penman-Monteith equation, which accounts for boundary layer effects. The details of `PhotosynEB` are not described here (see the documentation), because it is very similar to other implementations [@wang1998; @buckley2014].

@@

@@

# Acknowledgements {.unnumbered}
David Ellsworth, Belinda Medlyn, Martin de Kauwe are acknowledged for discussion on fitting A-Ci curves. Special thanks to Belinda Medlyn whose code in the MAESTRA model provided inspiration for an early version of the Photosyn function. Thanks to John Drake for testing.

# Figures {.unnumbered}

```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure1()
```

** Figure 1**. (a) Leaf photosynthesis - CO~2~ response curve as modelled with the FvCB model. (b) The intersection of the supply and demand curves of photosynthesis. The `Photosyn` function solves for C~i~ if g~s~, V~cmax~, J~max~ and R~d~ (and other parameters to the FvCB model) are known.

```{r fig.width=4, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure2()
```

**Figure 2**. Standard output from the `fitaci` function. A~n~ is the net photosynthetic rate, C~i~ the intercellular CO~2~ concentration. Symbols are measurements, the black line the fitted FvCB model of photosynthesis. Colored lines indicate the two photosynthesis rates in the FvCB model.

```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure3()
```

**Figure 3**. Response of A~n~ and C~i~ to combined changes in T~leaf~ and D. (a) Lines are A-C~i~ curves simulated at a range of values for T~leaf~. Symbols are the solutions of the coupled leaf gas exchange model, while also taking into account the correlation between D and T~leaf~ (based on an empirical relationship [@duursma2014]). Note that as T~leaf~ and D increase, C~i~ decreases. (b) The corresponding temperature optimum of A~n~. Symbols are the same as in panel (a) but plotted against T~leaf~.

```{r fig.width=8, fig.height=4, echo=FALSE, message=FALSE, eval=includeFigs, warn=FALSE}
figure4()
```

**Figure 4**. Visualization of the optimal model of stomatal conductance. Provided we have an estimate of the 'cost of water' ($\lambda$, mol C mol H~2~O^-1^), stomata act to maximize photosynthesis minus transpiration. In (a), individual curves at a range of values for the vapour pressure deficit (D) plots $A-\lambda*E$ versus the C~i~, demonstrating that an optimum exists. The ~FARAO~ function finds this optimum numerically and calculates A~n~ and g~s~. The corresponding response of g~s~ to D is shown in panel (b).

>>>>

*Short title*: The plantecophys R package

@@

# output
if(!dir.exists("output"))dir.create("output")

@@

# Set Flag to include figures in text or not.
includeFigs <- FALSE

@@

# Find current SHA of plantecophys
rep_pkg <- "c:/repos/plantecophys"
if(file.exists(rep_pkg)){
  pkgrep <- repository(rep_pkg)
  r <- reflog(pkgrep, verbose=FALSE)
  cur_sha <- substr(r[[1]]@sha, 1, 7)
} else {
  cur_sha <- "???"
}

# Make PDFs
to.pdf(figure1(), "output/figure1.pdf", width=8, height=4, verbose=FALSE)
to.pdf(figure2(), "output/figure2.pdf", width=4, height=4, verbose=FALSE)
to.pdf(figure3(), "output/figure3.pdf", width=8, height=4, verbose=FALSE)
to.pdf(figure4(), "output/figure4.pdf", width=8, height=4, verbose=FALSE)
```

@@

Keywords: photosynthesis, stomatal conductance, portable gas exchange instrument, process-based model, software

@@

Since the advent of portable gas exchange instruments [@mcdermitt1989; @leuning1989], a wealth of data on leaf gas exchange of CO~2~ and H~2~O has been collected  [@long2003]. These data play a central role in physiological plant ecology [@pearcey1989]. Both photosynthesis and transpiration vary widely between plant species, and fluctuate rapidly in response to environmental drivers such as light, humidity and temperature. Not only do leaf gas exchange data allow detailed studies of the underlying plant physiology, they are also used to parameterize an important component of land surface models used to predict global water and carbon cycling [e.g. @bonan2014; @dekauwe2015].

@@

\
(1)
$$A_n = min(A_c, A_j) - R_d$$

@@

\
(2)
$$C_c = C_i - A_n/g_m$$

@@

\
Fitting the model to data requires some finesse because the model is not a single non-linear equation, but instead a minimum function of two non-linear equations that is sometimes difficult to fit. Moreover, sample sizes collected are often small due to time constraints in the field. A widely used published method requires the user to specify the transition of V~cmax~ to J~max~ limitation [@sharkey2007], a process that is both arbitrary and prevents batch analysis. Another method [@leafweb] requires online submission of data and fits the model without much control or knowledge of the fitting process [following @gu2010]. Undoubtedly many more implementations of the fitting process have been developed over the years, but few of these are made publicly available (but see available online tools [@dekauwefitfarquhar; @landfluxtools]).

@@

\
(3)
$$ A_n = g_s(C_a - C_i) $$

\
where g~s~ is the conductance to CO~2~. We now have two equations for A~n~. The 'demand function' is given by Eq. 1, and the 'supply function' in Eq. 3. At steady state these two equation should be equal, which can be graphically shown as in Fig. 1 (cf. [@farquhar1982]).

\
**Figure 1**. (a) Leaf photosynthesis - CO~2~ response curve as modelled with the FvCB model. (b) The intersection of the supply and demand curves of photosynthesis. The `Photosyn` function solves for C~i~ if g~s~, V~cmax~, J~max~ and R~d~ (and other parameters to the FvCB model) are known.

\
Because g~s~ itself responds to environmental drivers, another expression is needed to end up with a fully coupled model of leaf gas exchange. The most successful, though empirical, g~s~ model is the Ball-Berry [@ballberry1987] class of models. This model posits an entirely empirical equation that describes the response of g~s~ to air humidity, CO~2~ and A~n~. This way, effects of leaf temperature and PPFD - both of which are known to affect g~s~ - are modelled through the dependency of A~n~ on these drivers. A general form of the Ball-Berry model is,

\
(4)
$$ g_s = g_0 + g_1 \frac{A_n}{C_a}f(D)$$

\
where D is the vapour pressure deficit (kPa), g~0~ and g~1~ are empirical parameters, and f(D) a function that describes the response to the vapour pressure deficit (D, [@leuning1995]) or relative humidity [@ballberry1987]. An alternative approach to modelling g~s~ is through the hypothesis that stomata act optimally in the sense that they maximize photosynthesis while minimizing water loss. This hypothesis was first developed by Cowan and Farquhar [@cowan1977] and has seen many applications. Medlyn et al. [@medlyn2011] showed that the optimality hypothesis, when coupled to the FvCB model, leads to an expression analogous to the Ball-Berry type models (Eq. 4), but with a different D response function (f(D) in Eq. 3) than the original Ball-Berry model.

\
Finally we can combine the biochemical demand function of photosynthesis (Eq. 1) with the supply function (Eq. 2) and an expression for the dependency of g~s~ on environmental drivers (Eq. 3). This 'coupled' leaf gas exchange model [@leuning1990; @collatz1991; @leuning1995] is implemented in many process-based ecosystem and global land surface models [@wang1998; @bonan2014; @duursma2012; @sellers1997]. This model allows prediction of A~n~, g~s~ and leaf transpiration rate in response to all major environmental drivers (except soil water limitation), and key leaf traits (g~1~, V~cmax~, J~max~, R~d~).

\
Despite the widespread use of the FvCB model and the coupled leaf gas exchange model, tools to analyze the data are scattered and subject to little standardization. Fitting the FvCB model to CO~2~ response curves is a standard procedure but different methods can yield different parameter values, making comparisons difficult. The coupled leaf gas exchange model is not straightforward to implement, and I do not know of any standalone open-source implementations. I here describe the plantecophys package, implemented in the R language [@rfoundation]. The code is freely available (without restrictions), and managed with a version control system. The package is the result of our work on leaf and canopy modelling of photosynthesis and stomatal conductance [@medlyn2011; @barton2012; @peltoniemi2012; @medlyn2013; @duursma2013; @duursma2014], with many additions based on user requests.

@@

The 'plantecophys' package is implemented in R, has no dependencies on other packages, and does not require compilation (i.e. it is written in native R only). As such it builds easily, and is highly portable. The source code is maintained with git version control, and is hosted in an online repository (http://www.bitbucket.org/remkoduursma/plantecophys). The repository includes an issue tracker, where users can suggest changes or report bugs.  This paper describes version `r packageVersion("plantecophys")` (git SHA `r cur_sha`).

@@

### Fitting A-C~i~ curves

The `fitaci` function fits the FvCB model, yielding estimates of V~cmax~, J~max~ and R~d~ and their standard errors. Optionally R~d~ can be provided as a known value, otherwise it is estimated from the A-C~i~ curve. The user does not have to provide the transition point (see Fig. 1), as this is estimated by `fitaci` automatically. It is however an option to fix the transition point (via the `citransition` argument), which may be helpful to check whether the best fit was achieved. Standard errors are estimated with the usual methods for non-linear regression (`nls` function in base R). Instead of fitting the minimum function (Eq. 1), `fitaci` uses the hyperbolic minimum of A~c~ and A~j~, which avoids a discontinuity. Finally, the user can provide an estimate of mesophyll conductance (g~m~) [following @ethier2004], in which case the fitted values of V~cmax~ and J~max~ can be interpreted as chloroplastic rates.

\
In the FvCB model, J~max~, V~cmax~ and leaf respiration (R~d~) (and other parameters like $\Gamma^{*}$, K~c~ and K~o~) all depend non-linearly on T~leaf~. The `Photosyn` function incorporates standard temperature sensitivities for all parameters of the FvCB model [following @medlyn2002].

@@

\
The output of the above example is shown in Fig. 2. Additionally, the batch utility `fitacis` can be used to fit many curves at once, for example one for each species or site in a dataset. The function `Aci` simulates the A-Ci curve, that is, it returns A~n~ when C~i~ is known.

\
**Figure 2**. Standard output from the `fitaci` function. A~n~ is the net photosynthetic rate, C~i~ the intercellular CO~2~ concentration. Symbols are measurements, the black line the fitted FvCB model of photosynthesis. Colored lines indicate the two photosynthesis rates in the FvCB model.

@@

The intersection of the supply and demand curves of photosynthesis (Fig. 1b) gives the operating intercellular CO~2~ concentration. This is solved by the `Photosyn` function. This flexible interface can be used to either 1) estimate A~n~ when C~i~ is known (`Photosyn(Ci=...)`, equivalent to `Aci(...)`), 2) estimate A~n~ when g~s~ is known (`Photosyn(GS=...)`) (cf. Fig. 1b) or c) solve for C~i~ from the coupled leaf gas exchange model (Eqs. 1,3,4).

\
To demonstrate the use of the coupled gas exchange model, I visualize the temperature response of A~n~ when both T~leaf~ and D are varying. In field conditions, D is always strongly positively related to T~leaf~. The consequence is that when studying D or T~leaf~ responses in the field, both drivers have to be accounted for [@lin2012; @duursma2014]. Figure 3a shows simulated A-C~i~ curves and the solutions of the coupled leaf gas exchange models at a range of T~leaf~ and corresponding D [calculated following @duursma2014]. Figure 3b shows the output from `Photosyn`, which can be created with the following example Note that in this example the default values of many parameters (e.g. J~max~, g~1~) are used in the call to `Photosyn`, but all of these can be set by the user.

\
**Figure 3**. Response of A~n~ and C~i~ to combined changes in T~leaf~ and D. (a) Lines are A-C~i~ curves simulated at a range of values for T~leaf~. Symbols are the solutions of the coupled leaf gas exchange model, while also taking into account the correlation between D and T~leaf~ (based on an empirical relationship [@duursma2014]). Note that as T~leaf~ and D increase, C~i~ decreases. (b) The corresponding temperature optimum of A~n~. Symbols are the same as in panel (a) but plotted against T~leaf~.

@@

\
The `Photosyn` function assumes that the boundary layer conductance (g~bl~) is high compared to g~s~, so that T~leaf~ is close to T~air~. As an alternative, the `PhotosynEB` function calculates T~leaf~ from the leaf energy balance. Transpiration is calculated with the Penman-Monteith equation, which accounts for boundary layer effects. The details of `PhotosynEB` are not described here (see the documentation), because it is very similar to other implementations [@wang1998; @buckley2014].

@@

\
**Figure 4**. Visualization of the optimal model of stomatal conductance. Provided we have an estimate of the 'cost of water' ($\lambda$, mol C mol H~2~O^-1^), stomata act to maximize photosynthesis minus transpiration. In (a), individual curves at a range of values for the vapour pressure deficit (D) plots $A-\lambda*E$ versus the C~i~, demonstrating that an optimum exists. The ~FARAO~ function finds this optimum numerically and calculates A~n~ and g~s~. The corresponding response of g~s~ to D is shown in panel (b).

@@

# Acknowledgments {.unnumbered}
David Ellsworth, Belinda Medlyn, Martin de Kauwe are acknowledged for discussion on fitting A-Ci curves. Special thanks to Belinda Medlyn whose code in the MAESTRA model provided inspiration for an early version of the Photosyn function. Thanks to John Drake for testing.
